Phenotypic evolution is more constrained in simpler food webs

### 2019-05-02 manuscript.Rmd
add manuscript changes on polynomial_model branch to master

\vspace*{0.1cm} 
\begin{center} \LARGE Phenotypic evolution is more constrained in simpler food webs \end{center}

@@

Global change is simplifying the structure of ecological networks; however, we are currently in a poor position to predict how these simplified communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web structure shape evolutionary constraints. We conducted a field experiment that manipulated the diversity of insect parasitoids that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that more traits were under selection in simpler vs. more complex food webs. The adaptive landscape was more neutral in complex food webs because different parasitoid species impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that the loss of species interactions can impose greater constraints on phenotypic evolution. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. As a control (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. To exclude the guild of larval parasitoids, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989]. In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. Since we were interested in selection imposed by parasitoids, we excluded unknown sources of mortality. For the food-web treatment that excluded larval parasitoids, we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

**Selection gradients**: We used the method of @Janzen1998 to translate trait-fitness relationships ($\alpha$) into the scale of relative fitness in order to calculate directional ($\beta_{z_i}$), quadratic ($\gamma_{z_i:z_i}$), and correlational ($\gamma_{z_i:z_j}$) selection gradients. Briefly, this method calculates the average gradient of selection surfaces by multiplying the average of $W(z)[1-W(z)]$ by each regression coefficient (e.g. $\alpha_{z_i}$, $\alpha_{z_i:z_i}$, or $\alpha_{z_i:z_j}$). We then divided this average gradient by mean fitness ($\bar W$) to put it on the scale of relative fitness, and thus interpretable as a selection gradient. We estimated selection gradients separately for each food-web treatment. Note that we doubled all quadratic terms prior to calculating selection gradients to put them on the same scale as estimates of directional and correlational selection [@Stinchcombe2008].

@@

## Measuring selection on the extended phenotype of egg parasitoids

Once parasitized, the gall phenotype becomes the extended phenotype of the egg parasitoid. This extended phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the extended phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify trait-fitness relationships and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web structure on the egg parasitoid's adaptive landscape ---we can only estimate the selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the community context may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Loss of species interactions increase selective constraints

@@

## Loss of species interactions increase genetic constraints

@@

## Selection on the extended phenotype of egg parasitoids

@@

<!--Summary-->
We found that the extinction of larval parasitoids constrained phenotypic evolution in gall midges in two key ways. First, more traits contributed to the slope of the adaptive landscape in the absence of larval parasitoids, suggesting greater constraints on the trajectory of phenotypic evolution. Second, excluding larval parasitoids indirectly increased the number of genetic constraints, which could act to constrain the gall midge's adaptive potential in the face of novel selection pressures. Our experiment also revealed evidence of indirect selection pressures, suggesting that the loss of species interactions may have complex effects on the trajectories of phenotypic evolution. Taken together, our study provides experimental evidence from the field that the loss of species interactions may constrain the adaptive potential of remaining populations.

@@

<!--Indirect evolutionary effects-->
Our results suggest that the loss of species interactions may not only directly affect connected species, but also result in indirect evolutionary effects. In our study, this indirect effect arises from egg parasitoids being released from intraguild predation when we exlcuded larval parasitoids. This release occurs more on trees with high larval densities, which could intensify selection on gall midge oviposition preference. A growing number of experiments over the past two decades have demonstrated the presence and potential importance of indirect evolutionary effects that emerge in a community context [@Juenger1998; @Stinchcombe2001; @Lankau2007; @Walsh2008; @Walsh2010; @terHorst2010; @Sahli2011; @Lau2012; @terHorst2015; @Schiestl2018; @StartInPress]. If indirect evolutionary effects are common [@Walsh2013], then predicting evolutionary trajectories resulting from the loss of species interactions will require evolutionary studies to explicitly account for the ecological networks that species are embedded in.

>>>>

\vspace*{0.1cm} 
\begin{center} \LARGE Extinction of a consumer guild constrains phenotypic evolution in an insect herbivore \end{center}

@@

Global change is altering the structure of ecological networks; however, we are currently in a poor position to predict how these altered communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web structure shape evolutionary constraints. We conducted a field experiment that manipulated whether a guild of larval parasitoids were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that more traits were under selection when we excluded interactions with larval parasitoids. The adaptive landscape was more neutral in the intact food web because different parasitoid guilds impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that the loss of trophic interactions can impose greater constraints on phenotypic evolution. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. As a control (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. To exclude the guild of larval parasitoids, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989].
In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. Since we were interested in selection imposed by parasitoids, we excluded other sources of mortality (`r round((mean(full_df$early.larval.death) + mean(full_df$unknown_mortality)), 2)*100`%).

For the food-web treatment that excluded larval parasitoids, we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

**Selection gradients**: Selection gradients cannot be estimated directly from generalized linear (mixed) models, because the coefficients are on a nonlinear scale. For example, the coefficients in the above model measure the change in the logarithm of the odds of surviving (i.e., $\text{ln}\{W(z)[1-W(z)]\}$) associated with a unit change in a trait with all other traits held fixed. Therefore, we used the method developed by @Janzen1998 to translate trait-fitness relationships from the above model into the scale of relative fitness in order to calculate directional ($\beta_{z_i}$), quadratic ($\gamma_{z_i:z_i}$), and correlational ($\gamma_{z_i:z_j}$) selection gradients. Briefly, this method calculates the average gradient of selection surfaces by multiplying the average of $W(z)[1-W(z)]$ by each regression coefficient (e.g. $\alpha_{z_i}$, $\alpha_{z_i:z_i}$, or $\alpha_{z_i:z_j}$). We then divided this average gradient by mean fitness ($\bar W$) to put it on the scale of relative fitness, and thus interpretable as a selection gradient. We estimated selection gradients separately for each food-web treatment. Note that we doubled all quadratic terms prior to calculating selection gradients to put them on the same scale as estimates of directional and correlational selection [@Stinchcombe2008].

@@

## Measuring selection on egg parasitoids

Once parasitized, the gall phenotype becomes the phenotype of the egg parasitoid. This phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify trait-fitness relationships and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web structure on the egg parasitoid's adaptive landscape ---we can only estimate the selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the community context may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Loss of trophic interactions increase selective constraints

@@

## Loss of trophic interactions increase genetic constraints

@@

## Selection on egg parasitoids

@@

<!--Summary-->
We found that the extinction of larval parasitoids constrained phenotypic evolution in gall midges in two key ways. First, more traits contributed to the slope of the adaptive landscape in the absence of larval parasitoids, suggesting greater constraints on the trajectory of phenotypic evolution. Second, excluding larval parasitoids indirectly increased the number of genetic constraints, which could act to constrain the gall midge's adaptive potential in the face of novel selection pressures. Our experiment also revealed evidence of indirect selection pressures, suggesting that the loss of species interactions may have complex effects on the trajectories of phenotypic evolution. Taken together, our study provides experimental evidence from the field that the loss of trophic interactions may constrain the adaptive potential of remaining populations.

@@

<!--Indirect evolutionary effects-->
Our results suggest that the loss of trophic interactions may not only directly affect connected species, but also result in indirect evolutionary effects. In our study, this indirect effect arises from egg parasitoids being released from intraguild predation when we exlcuded larval parasitoids. This release occurs more on trees with high larval densities, which could intensify selection on gall midge oviposition preference. A growing number of experiments over the past two decades have demonstrated the presence and potential importance of indirect evolutionary effects that emerge in a community context [@Juenger1998; @Stinchcombe2001; @Lankau2007; @Walsh2008; @Walsh2010; @terHorst2010; @Sahli2011; @Lau2012; @terHorst2015; @Schiestl2018; @StartInPress]. If indirect evolutionary effects are common [@Walsh2013], then predicting evolutionary trajectories resulting from the loss of species interactions will require evolutionary studies to explicitly account for the ecological networks that species are embedded in.
